\chapter{Conclusions and Outlook}
\label{chap:conclusions}

The Standard Model of particle physics provides the most precise description of 
fundamental physics and remains  the most experimentally verified
model available. The mechanism by which electroweak symmetry breaking occurs 
in the standard model, giving rise to the masses of the fundamental fermions and
bosons, predicts the existence of a new massive scalar boson, the Higgs boson.
Such a particle should be experimentally observable, although prior to the LHC being
turned on, no such particle had been discovered.

In this thesis, a search for this particle, through its decay to two photons, in proton-proton collisions recorded at the CMS detector, during 2011 at a centre of mass $\sqrt{s}=7$ TeV
has been described. 
The decay channel, despite having a 
relatively low branching ratio, is one of the most sensitive at CMS 
due to the high resolution 
of the electromagnetic calorimeter and the narrow invariant mass peak it provides.
The analysis detailed employed the use of several multivariate analysis techniques
in order to provide the greatest sensitivity to a potential signal.
As the signal yield in the two photon decay channel is small, 
the search for $\Hgg$ is highly sensitive to the background modelling.
The signal extraction technique described in this thesis was one which was 
developed by the author and served as a cross-check of the published result from the 2011 dataset.
This allowed for additional scrutiny on the background modelling 
to which the search in this decay channel is so sensitive. When combined with additional 
data from 2012 at a centre of mass energy $\sqrt{s}=8$ TeV, an excess near 125 GeV 
was observed with a significance of around $4\sigma$.

In order to maximise the sensitivity of the search for the Standard Model Higgs boson,
data from several decay channels are combined at CMS using the methods described in this thesis. The limits on Standard Model Higgs boson production were the most stringent at the 
time of the International Conference of High Energy Physics in 2012.
An excess was observed in the combined 2011 and 2012 datasets which is 
compatible with a Standard Model Higgs boson with a mass of 125 GeV. The excess was 
significant enough so as to claim discovery at the $5\sigma$ level. The excess is driven 
by the $\Hgg$ and  $\Hzzl$ channels, although additional evidence is found in the $\Hww$ channels. With the data available by the Hadron Collider Physics symposium of November 2012,
study of its couplings to Standard Model particles indicates that the new particle is 
consistent with the Standard Model Higgs boson, though additional data are required to 
make a definitive statement.

The discovery of the new particle is one of great significance to particle physics.
Should the particle turn out to conform to the predictions of the Standard Model, 
its discovery will have provided a great step into understanding the nature of 
electroweak symmetry breaking. However, if this turns out not to be the case, 
deviations from the predictions will indicate hints of potential new physics
and serve as guidance in the search for physics beyond the Standard Model . 
Additional data will be taken once the LHC resumes 
collisions in 2015 with an increased centre-of-mass energy, $\sqrt{s}=14$ TeV. 
With the additional data, stronger statements can be made as to 
the exact nature of the new particle and its interactions with the Standard Model or physics beyond it. With this discovery in hand and the search potentially at an end, 
it is clear that a new window into fundamental physics has been opened, and the intriguing 
studies into what that physics is has only now begun.




